% !Mode:: "TeX:UTF-8"

\cabstract{
针对管道输送系统中泄漏状态识别困难、人工特征设计依赖较强以及泄漏距离估计稳定性不足等问题，本文围绕双通道时序传感信号，研究一种两阶段管道泄漏检测与定位方法。首先，结合泄漏声信号形成机制与双通道传播差异，构建从原始 CSV 采样数据到训练样本的自动化处理流程，将长时序双通道信号按照固定时间窗口切分为分段样本，并分别生成分类任务与回归任务所需的数据清单。其次，面向泄漏状态识别与泄漏距离估计两个目标，设计统一训练框架。本文沿用项目配置命名：Stage2 表示分类任务，以单通道分段信号为输入，用于识别正常状态与多类泄漏形态；Stage1 表示距离回归任务，以左右双通道信号对为输入，用于估计泄漏距离。模型方面，本文以一维卷积网络作为基础结构，并补充 Conformer 结构、传统机器学习基线和通道消融实验，以分析不同建模方式在分类与回归任务中的适用边界。实验结果表明，基础一维卷积模型在 Stage2 测试集上取得片段级 Accuracy 0.9415、Macro-F1 0.9180，文件级 Accuracy 0.9783、Macro-F1 0.9836；Conformer 进一步将片段级 Accuracy 提升至 0.9900、Macro-F1 提升至 0.9883，文件级两项指标均达到 1.0000。在 Stage1 距离回归任务中，基础一维卷积模型取得片段级 MAE 1.90 m、RMSE 2.43 m，验证了从双通道原始波形中学习距离映射的可行性，但 SVM 基线和 Conformer 对比结果也表明，当前回归定位仍受样本规模、距离档位分布和细粒度信号差异制约。上述结果说明，两阶段框架能够有效支持泄漏识别与定位分析，同时后续仍需在数据覆盖、距离回归建模和跨工况泛化验证方面进一步改进。
}

\ckeywords{管道泄漏检测，时序信号分析，一维卷积神经网络，Conformer，距离回归}

\eabstract{
To address the difficulty of leak state recognition, the dependence on handcrafted features, and the limited stability of leak distance estimation in pipeline transport systems, this thesis studies a two-stage pipeline leak detection and localization method based on dual-channel sensor data. First, considering the generation mechanism of leakage acoustic signals and the propagation differences between two sensing channels, an automated data processing pipeline is constructed from raw CSV measurements. Long dual-channel signals are segmented with a fixed time window to generate structured samples for classification and regression tasks. Second, a unified training framework is designed for leak state recognition and distance estimation. Following the project configuration, Stage2 denotes the classification task, where segmented single-channel signals are used to identify normal states and multiple leak patterns; Stage1 denotes the distance regression task, where paired left-right channel signals are used to estimate leak distance. In terms of model design, a one-dimensional convolutional neural network is used as the basic architecture, while Conformer models, traditional machine-learning baselines, and channel ablation experiments are further introduced to analyze the applicability of different modeling strategies. Experimental results show that the basic one-dimensional convolutional model achieves a segment-level Accuracy of 0.9415 and Macro-F1 of 0.9180 on Stage2, with file-level Accuracy and Macro-F1 reaching 0.9783 and 0.9836, respectively. The Conformer model further improves the segment-level Accuracy to 0.9900 and Macro-F1 to 0.9883, and achieves 1.0000 on both file-level metrics. On the Stage1 distance regression task, the basic convolutional model obtains a segment-level MAE of 1.90 m and RMSE of 2.43 m, demonstrating the feasibility of learning distance-related mappings from raw dual-channel waveforms. However, the SVM baseline and Conformer comparison also indicate that distance regression is still constrained by sample size, distance-bin distribution, and fine-grained signal differences. These results show that the two-stage framework can support leak recognition and localization analysis, while further improvements are still required in data coverage, regression modeling, and cross-condition generalization.
}

\ekeywords{Leak Detection; Time-series; 1D CNN; Conformer; Regression}

\makeabstract
